---
title: "Working with NEON data on mobile devices"
author: "Matthew Aiello-Lammens"
output: html_document
---

```{r, eval=FALSE}
install.packages("devtools")
install.packages("dplyr")
install.packages("neonUtilities")
devtools::install_github("NEONScience/NEON-geolocation/geoNEON")
```

Load packages

```{r}
library(neonUtilities)
library(dplyr)
library(ggplot2)
#library(geoNEON)
#library(raster)
#library(rhdf5)

# Set global option to NOT convert all character variables to factors
options(stringsAsFactors=F)
```


# Working with plant presence and percent cover data

We will begin by working with NEONs **plant presence and percent cover** data set. 
More details on these data can be found here - [https://data.neonscience.org/data-products/DP1.10058.001](https://data.neonscience.org/data-products/DP1.10058.001).

## Getting data using `loadByProduct`

First, we will use the `loadByProduct` function in `neonUtilities`.

**NOTE:** Here we are using the `check.size = FALSE` argument. 
We are doing this to prevent the function from sending a query to the console, and because we know these data are relatively small in size (approx. 2.5 MB). 
However, if working with *potentially* larger data sets, then you should set `check.size = TRUE`.

```{r}
plants <- loadByProduct(dpID = "DP1.10058.001", site = c("HARV", "BART"), check.size = FALSE)
```

## Alternative approach - working with zipped files after manual download from data portal

Stack data

```{r, eval = FALSE}
stackByTable("NEON_presence-cover-plant.zip")
```

Read data

```{r, eval = FALSE}
plant_data_1m <- read.csv("NEON_presence-cover-plant/stackedFiles/div_1m2Data.csv")
```


## Exploring the data

Let's examine the data we just recieved. They should be stored as a `list` object. 
You can find out more information about lists here - [https://stat.ethz.ch/R-manual/R-devel/library/base/html/list.html](https://stat.ethz.ch/R-manual/R-devel/library/base/html/list.html).

For our immediate perposes, what we need to know is that we can access any element of a `list` with the `$` operator. But first, let's get the names of these different objects.

```{r}
names(plants)
```

The `readme_10058` element looks promising.

```{r}
plants$readme_10058
```

OK. If we navigated that readme just a bit, we would have learned that there are at least two data sets with the actual vegetation survey data: `div_1m2Data` and `div_10m2Data_100m2Data`. 

Let's have a peek at each.

```{r}
head(plants$div_1m2Data)
```


```{r}
head(plants$div_10m2Data100m2Data)
```

How many rows are in each?

1m2 plots

```{r}
nrow(plants$div_1m2Data)
```

10m2 / 100m2 plots

```{r}
nrow(plants$div_10m2Data100m2Data)
```

Those are tens of thousands of rows!

Let's have a look at how many unique species occur in each data set.

1m2 plots

```{r}
length(unique(plants$div_1m2Data$scientificName))
```

10m2 / 100m2 plots

```{r}
length(unique(plants$div_10m2Data100m2Data$scientificName))
```

## Calculating data summaries

We are going to do some quick and dirty data summaries.

First, let's calculate species richness values for plots.
Here, we are going to work at the **plot** level, including all of the data from the subplots.
To do this, we first need to make a new column that describes just the plot number.


Combine data sets

```{r}
all_plants <- rbind(
  select(plants$div_10m2Data100m2Data, siteID, plotID, subplotID, endDate, scientificName, family, nativeStatusCode), 
  select(plants$div_1m2Data, siteID, plotID, subplotID, endDate, scientificName, family, nativeStatusCode))

all_plants <- filter(all_plants, !is.na(scientificName))
```



```{r}
all_plants$mainsubplotID <- as.numeric(gsub(pattern = "\\..*", replacement = "", all_plants$subplotID))
```



```{r}
all_plants_summary <-
  all_plants %>%
  group_by(siteID, plotID) %>%
  summarise(Richness = length(unique(scientificName)))
all_plants_summary
```


```{r}
ggplot(data = all_plants_summary) +
  geom_histogram(aes(x = Richness, fill = siteID), position = "dodge")
```



```{r}
length(unique(all_plants$plotID))
length(unique(all_plants$mainsubplotID))
```





```{r}
plant_data %>%
  filter(scientificName == "Frangula alnus Mill.")


plant_data %>%
  group_by(plotID) %>%
  summarise(richness = length(unique(scientificName)))
```

