---
title: "Working with NEON data on mobile devices"
author: "Matthew Aiello-Lammens"
output: html_document
---

# Overview

In this activity, you will be introduced to some basic data processing and exploration steps using data from the National Ecological Observatory Network (NEON). 
You can learn more about NEON at this website - [https://www.neonscience.org/](https://www.neonscience.org/).
This activity is designed so that you can complete it using a mobile device, such as your phone, in conjunction with RStudio Cloud, [https://rstudio.cloud/](https://rstudio.cloud/).
Throughout the first part of the activity, we will work with the NEON **Plant Presence and Percent Cover** data set. 
More details on these data can be found here - [https://data.neonscience.org/data-products/DP1.10058.001](https://data.neonscience.org/data-products/DP1.10058.001).

# Learning Objectives

By completing this activity, you should be able to:

## Quantitative / Data Skills

* explain the importance of data management in the scientific process
* modify a data set using a `data.frame` object in R
* modify `dplyr` code to summerise a data set
* calculate plot-level species richness values

## Concept learning objectives

* interpret data visualizations, such as histograms
* develop hypotheses related to differences in plant species richness among NEON sites



# Preliminaries

## Starting your RStudio session

In order to carry out the analyses in this activity, you will first need to initate a new R session. 
If you are on a desktop computer with RStudio installed, start a new RStudio session. 
(**NOTE: In this activity, I am not being careful in distiguishing R and RStudio. Strictly speaking, this is poor practice. R is the computer programming language and RStudio is an Integrated Development Environment that makes using R easier. You need to have both programs installed on your desktop.**)
If you do not have RStudio installed, or you are working on a mobile device, you can sign-up for a free RStudio Cloud account, where you can run an RStudio session through a browser. 
Go to [https://rstudio.cloud/](https://rstudio.cloud/) to begin this process.

## Getting our Rmd file 

Next, let's get the Rmd file that will allow us to run all of the code for this activity.
This Rmd file is also the file that translates into this webpage! (Yes, that's meta.)
We will use the `download.file` function to get this Rmd file.
You will need to copy and paste this line in your RStudio session.

```{r, eval=FALSE}
download.file("https://github.com/mlammens/NEONonMobile/raw/master/working-with-neon.Rmd",
              destfile = "./temp.Rmd")
```



## Install and load a few packages into your R environment

If you are working on a mobile device ...


```{r, eval=FALSE}
install.packages("dplyr")
install.packages("ggplot2")
install.packages("neonUtilities")
```


Load packages

```{r}
library(neonUtilities)
library(dplyr)
library(ggplot2)

# Set global option to NOT convert all character variables to factors
options(stringsAsFactors=F)
```


# Working with plant presence and percent cover data

We will begin by working with NEONs **plant presence and percent cover** data set. 
More details on these data can be found here - [https://data.neonscience.org/data-products/DP1.10058.001](https://data.neonscience.org/data-products/DP1.10058.001).

## Getting data using `loadByProduct`

First, we will use the `loadByProduct` function in `neonUtilities`.

**NOTE:** Here we are using the `check.size = FALSE` argument. 
We are doing this to prevent the function from sending a query to the console, and because we know these data are relatively small in size (approx. 2.5 MB). 
However, if working with *potentially* larger data sets, then you should set `check.size = TRUE`.

```{r}
plants <- loadByProduct(dpID = "DP1.10058.001", site = c("HARV", "BART"), check.size = FALSE)
```

## Alternative approach - working with zipped files after manual download from data portal

Stack data

```{r, eval = FALSE}
stackByTable("NEON_presence-cover-plant.zip")
```

Read data

```{r, eval = FALSE}
plant_data_1m <- read.csv("NEON_presence-cover-plant/stackedFiles/div_1m2Data.csv")
```


## Exploring the data

Let's examine the data we just recieved. They should be stored as a `list` object. 
You can find out more information about lists here - [https://stat.ethz.ch/R-manual/R-devel/library/base/html/list.html](https://stat.ethz.ch/R-manual/R-devel/library/base/html/list.html).

For our immediate perposes, what we need to know is that we can access any element of a `list` with the `$` operator. But first, let's get the names of these different objects.

```{r}
names(plants)
```

The `readme_10058` element looks promising.

```{r}
plants$readme_10058
```

OK. If we navigated that readme just a bit, we would have learned that there are at least two data sets with the actual vegetation survey data: `div_1m2Data` and `div_10m2Data_100m2Data`. 

Let's have a peek at each.

```{r}
head(plants$div_1m2Data)
```


```{r}
head(plants$div_10m2Data100m2Data)
```

How many rows are in each?

1m2 plots

```{r}
nrow(plants$div_1m2Data)
```

10m2 / 100m2 plots

```{r}
nrow(plants$div_10m2Data100m2Data)
```

Those are tens of thousands of rows!

Let's have a look at how many unique species occur in each data set.

1m2 plots

```{r}
length(unique(plants$div_1m2Data$scientificName))
```

10m2 / 100m2 plots

```{r}
length(unique(plants$div_10m2Data100m2Data$scientificName))
```

## Calculating data summaries

We are going to do some quick and dirty data summaries.

First, let's calculate species richness values for plots.
Here, we are going to work at the **plot** level, including all of the data from the subplots.
To do this, we first need to make a new column that describes just the plot number.


Combine data sets

```{r}
all_plants <- rbind(
  select(plants$div_10m2Data100m2Data, siteID, plotID, subplotID, endDate, scientificName, family, nativeStatusCode), 
  select(plants$div_1m2Data, siteID, plotID, subplotID, endDate, scientificName, family, nativeStatusCode))

all_plants <- filter(all_plants, !is.na(scientificName))
```



```{r}
all_plants$mainsubplotID <- as.numeric(gsub(pattern = "\\..*", replacement = "", all_plants$subplotID))
```



```{r}
all_plants_summary <-
  all_plants %>%
  group_by(siteID, plotID) %>%
  summarise(Richness = length(unique(scientificName)))
all_plants_summary
```


```{r}
ggplot(data = all_plants_summary) +
  geom_histogram(aes(x = Richness, fill = siteID), position = "dodge")
```


```{r}
t.test(all_plants_summary$Richness ~ all_plants_summary$siteID)
```



```{r}
length(unique(all_plants$plotID))
length(unique(all_plants$mainsubplotID))
```
